Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory Concentrations

### 2015-06-04 manuscript/manuscript.Rmd
updated with chla in abstract
updates.  still needs to be rendered.

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide arrange of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect agains the impacts, several health advisory levels have been set for some of these toxins, in particular, one of the most common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, measuring micorcystin can not currently be done *in situ* and requires samples be processed in a lab.  This time consuming and expensive.  It is possible to find reliable indicators that may be estimated quickly and *in situ* as a first defense agains high level of microcystin.  In particular, chlorophyll *a* has been shown to be postively associated with microcystin.  In this paper we use this assocation to provide estimates of chlorophyll *a* that if exceeded would be indiciative of a higher likelihood of exceeding select concentrations of microcystin.  Using the 2007 National Lakes Assessment and a conditional probability appoach that has been used in other water quality settings, we idenfify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recerational use.  For microcystin concentrations of 0.3, 1, 1.6, 2. and 4 we find chlorophyll *a* concentrations of [R CODE HERE] respectively.  When managing for these various microsystin levels exceed these reproted chlorophyll *a* concentratoins should be a trigger for further testing and possibly managment action.

@@

EPA just released HA for MC

>>>>

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide arrange of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect agains the impacts, several health advisory levels have been set for some of these toxins, in particular, one of the most common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, measuring micorcystin can not currently be done *in situ* and requires samples be processed in a lab.  This time consuming and expensive.  It is possible to find reliable indicators that may be estimated quickly and *in situ* as a first defense agains high level of microcystin.  In particular, chlorophyll *a* has been shown to be postively associated with microcystin.  In this paper we use this assocation to provide estimates of chlorophyll *a* that if exceeded would be indiciative of a higher likelihood of exceeding select concentrations of microcystin.  Using the 2007 National Lakes Assessment and a conditional probability appoach that has been used in other water quality settings, we idenfify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recerational use.  For microcystin concentrations of 0.3, 1, 1.6, 2. and 4 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, `r mc_chla[4]`, and `r mc_chla[5]`, respectively.  When managing for these various microsystin levels exceed these reproted chlorophyll *a* concentratoins should be a trigger for further testing and possibly managment action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down their municipal water supply due in part to an excess of Microcystin-LR that resulted from a ongoin harmful algal bloom in Lake Erie.  Since this event, signficant legislation has been passed in the United States and the US Environmental Protection Agency has released suggested microcystin-LR concentrations that would trigger health advisories.
